---
title: "moorea_heatwave"
author: "daniellembecker"
date: "1/15/2022"
output:
  pdf_document: default
  html_document: default
adapted from: amandachiachi
---

#OISST_data
#Adapted by Amanda Chiachi
#11/13/2019

#### Setup ##########

#load libraries
```{r}
library(dplyr)
library(ggplot2)
library(heatwaveR)# The packages we will use
library(dplyr) # A staple for modern data management in R
library(lubridate) # Useful functions for dealing with dates
library(ggplot2) # The preferred library for data visualisation
library(tidync) # For easily dealing with NetCDF data
library(rerddap) # For easily downloading subsets of data
library(doParallel) # For parallel processing
library(lubridate)
library(scales)
library(maps)

```



# The information for the NOAA OISST data
```{r}
rerddap::info(datasetid = "ncdcOisst21Agg_LonPM180", url = "https://coastwatch.pfeg.noaa.gov/erddap/")
```


# This function expects the user to provide it with a start and end date
# It then downloads and prepares the data
# we are downloading only the SST data out of several variables 
# spatial extent of latitude -17.466987, -17.597282 & longitude -149.849395, -149.822959
```{r}
OISST_sub_dl <- function(time_df){
  OISST_dat <- griddap(x = "ncdcOisst21Agg_LonPM180", 
                       url = "https://coastwatch.pfeg.noaa.gov/erddap/", 
                       time = c(time_df$start, time_df$end), 
                       zlev = c(0, 0),
                       latitude = c(-17.466987, -17.597282),
                       longitude = c(-149.849395, -149.822959),
                       fields = "sst")$data %>% 
    mutate(time = as.Date(stringr::str_remove(time, "T00:00:00Z"))) %>% 
    dplyr::rename(t = time, temp = sst) %>% 
    select(lon, lat, t, temp) %>% 
    na.omit()
}

```




#server doesnt like more than 9 years of consecutive data 
#creates a data frame to download multiple batches of the data 
# allows us to automate the entire download 
# Date download range by start and end dates per year
```{r}
moorea_years <- data.frame(date_index = 1,
                                 start = as.Date(c("2002-01-01")),
                                 end = as.Date(c("2021-11-01")))
                    

system.time(
  OISST_data <- moorea_years %>% 
    group_by(date_index) %>% 
    group_modify(~OISST_sub_dl(.x)) %>% 
    ungroup() %>% 
    select(lon, lat, t, temp)
)


#format date and time, lubridate package format
OISST_data$t <- ymd(OISST_data$t) 

#subset temp data for March through April for 2019 and 2020

heatwave.temps.OISST <- OISST_data %>%
  filter(lubridate::month(t) %in% c(3:4)) 

heatwave.temps_filter.OISST <- heatwave.temps.OISST %>%
  filter(lubridate::year(t) %in% c(2019:2021)) 

write.csv(heatwave.temps_filter.OISST, "../../RAnalysis/output/heatwave/moorea.heatwave.temps.OISST.csv")

```



# Detect the events in a time series
# "...must be at least three years to calculate thresholds"
# create your time series data 
```{r}
ts.OISST <- ts2clm(OISST_data, climatologyPeriod = c("2002-01-01", "2021-11-01"))
mhw.OISST <- detect_event(ts.OISST)
```


# View just a few metrics
# The function will return a line plot indicating the climatology, threshold and temperature, with the hot or cold events that meet the specifications of Hobday et al. (2016) shaded in as appropriate.
# An anomalously warm event is considered to be a MHW if it lasts for five or more days, with temperatures warmer than the 90th percentile based on a 30-year historical baseline period.
```{r}
#make data frame with marine heatwave events over time in Mo'orea

mhw.record.OISST <- mhw.OISST$event %>% 
  dplyr::ungroup() %>%
  dplyr::select(event_no, duration, date_start, date_end, date_peak, intensity_max, intensity_cumulative, intensity_mean, rate_onset, rate_decline) %>% 
  dplyr::arrange(-intensity_max); mhw.record.OISST 

write.csv(mhw.record.OISST, "RAnalysis/output/heatwave/marine.heatwave.record.OISST.csv")

#format date and time
mhw.OISST$Date<- mdy(mhw.OISST$Date) 


#create a line plot of marine heatwave data, can change date range for more specific times
#2019 heatwave scenarios
event_line(mhw.OISST, spread = 50, metric = "intensity_max", 
           start_date = "2019-03-01", end_date = "2019-05-01", x_axis_title = "Date") +
    scale_x_date(date_breaks = "1 day", labels = date_format("%b-%d")) + 
    theme_classic()+
    theme(axis.text.x = element_text(angle = 90)) 


```

```{r}
#2020 heatwave scenarios
event_line(mhw.OISST, spread = 50, metric = "intensity_max", 
           start_date = "2020-03-01", end_date = "2020-05-01", x_axis_title = "Date") +
    scale_x_date(date_breaks = "2 days", labels = date_format("%b-%d")) + 
    theme_classic()+
    theme(axis.text.x = element_text(angle = 90)) 

```

```{r}
#2021 heatwave scenarios
event_line(mhw.OISST, spread = 50, metric = "intensity_max", 
           start_date = "2021-03-01", end_date = "2021-05-01", x_axis_title = "Date") +
    scale_x_date(date_breaks = "2 days", labels = date_format("%b-%d")) + 
    theme_classic()+
    theme(axis.text.x = element_text(angle = 90)) 
```



####Use heatwave package on LTER 02 backreef temperature data

#format LTER2 dataframe to use for mhw scenario
```{r}

#load in LTER 02 backreef data from 2005 to 2021, delete first row
LTER2 <- read.csv("RAnalysis/output/temperature/LTER_2_temp_backreef.csv", row.names = 1)

#remove columns not needed for analysis (site,time,reef.type,and depth)

LTER2.mhw <- LTER2[ -c(1,3:5) ]

#rename date column to t

LTER2.mhw <- rename(LTER2.mhw, t = Date)

#need to get data point per day to match OISST processing

LTER2.mhw.dat <- LTER2.mhw %>%
  group_by(t) %>% #tells to group by treatment
  summarise(mean=mean(temp, na.rm=TRUE)) #calculates mean and se
LTER2.mhw.dat

#add lat and lon columns to data frame 

LTER2.mhw <- LTER2.mhw.dat %>%
  mutate(lon = -149.8116849,)

LTER2.mhw <- LTER2.mhw %>%
  mutate(lat = -17.48131958)

#rearrange columns in order for mhw scenario

LTER2.mhw.reorder <- LTER2.mhw[, c(3,4, 1, 2)]

LTER2.mhw.reorder

#rename mean column to temp

LTER2.mhw <- rename(LTER2.mhw.reorder, temp = mean)

#format date and time, lubridate package format
LTER2.mhw$t <- ymd(LTER2.mhw$t) 

LTER2.mhw


#subset temp data for March through April for 2019 and 2020

heatwave.temps.lter <- LTER2.mhw %>%
  filter(lubridate::month(t) %in% c(3:4)) 

heatwave.temps_filter.lter <- heatwave.temps.lter %>%
  filter(lubridate::year(t) %in% c(2019:2021)) 

write.csv(heatwave.temps_filter.lter, "RAnalysis/output/heatwave//moorea.heatwave.temps.lter.csv")

```

# Detect the events in a time series
# create your time series data 
```{r}
ts.lter <- ts2clm(LTER2.mhw, climatologyPeriod = c("2005-06-01", "2021-07-06"))
mhw.lter <- detect_event(ts.lter)

```


# View just a few metrics
# The function will return a line plot indicating the climatology, threshold and temperature, with the hot or cold events that meet the specifications of Hobday et al. (2016) shaded in as appropriate.
# An anomalously warm event is considered to be a MHW if it lasts for five or more days, with temperatures warmer than the 90th percentile based on a 30-year historical baseline period.
```{r}
#make data frame with marine heatwave events over time in Mo'orea

mhw.record.lter <- mhw.lter$event %>% 
  dplyr::ungroup() %>%
  dplyr::select(event_no, duration, date_start, date_end, date_peak, intensity_max, intensity_cumulative, intensity_mean, rate_onset, rate_decline) %>% 
  dplyr::arrange(-intensity_max); mhw.record.lter 

write.csv(mhw.record.lter, "RAnalysis/output/heatwave/marine.heatwave.record.lter.csv")



#create a line plot of marine heatwave data, can change date range for more specific times
#2019 heatwave scenarios
event_line(mhw.lter, spread = 30, metric = "intensity_max", 
           start_date = "2019-03-30", end_date = "2019-04-10", x_axis_title = "Date") +
    scale_x_date(date_breaks = "1 day", labels = date_format("%b-%d")) + 
    theme_classic()+
    theme(axis.text.x = element_text(angle = 90)) 

```
```{r}

#2020 heatwave scenarios
event_line(mhw.lter, spread = 50, metric = "intensity_max", 
           start_date = "2020-03-01", end_date = "2020-05-01", x_axis_title = "Date") +
    scale_x_date(date_breaks = "2 days", labels = date_format("%b-%d")) + 
    theme_classic()+
    theme(axis.text.x = element_text(angle = 90)) 
```

```{r}

#make dataframe and plot for sampling process

sampling.dat <- read.csv("RAnalysis/data/temperature/sampling.plan.data.csv")

#format date for lubridate
sampling.dat$Date <- mdy(sampling.dat$Date) 

#plot mean temp per day by date
sampling.plot <- sampling.dat %>%
  ggplot(aes(x=Date, y=Temperature, color = Sampling))+
  geom_line() + 
  geom_point() +
  scale_x_date(date_breaks = "1 day", date_labels = "%m-%d") +
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90)) +
  ylab("Temperature (°C)")+
  xlab("Date");sampling.plot


ggsave(filename="RAnalysis/output/heatwave/sampling.plot.pdf", plot=sampling.plot, width=10, height=10, units="in")
```



        
