---
title: "manuscript_figures"
author: "daniellembecker"
date: "2024-05-23"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Use this script to use cleaned data and make plots for RI kelp data manuscript

# Load libraries 
```{r}
library(plyr)
library(dplyr)
library(tigris)
library(car)
library(MuMIn)
library(lmerTest)
library(tidyverse)
library(lme4)
library(broom.mixed)
library(ggmap)
library(lubridate)
library(ggplot2)
library(patchwork)
library(broom)
library(ggspatial)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggspatial)

```

# load cleaned dataframes

```{r}
# Set your directory path
directory <- "output/cleaned_data/"

# List files in the directory
files <- list.files(directory)

# Initialize an empty list to store data frames
data_frames <- list()

# Loop through each file
for (file in files) {
    # Read the file into a data frame
    # Adjust the read function according to your file type (e.g., read.csv, read.table)
    # Specify other parameters as needed (e.g., header = TRUE)
    file_path <- file.path(directory, file)
    df_name <- gsub("\\..*", "", file)  # Extract file name without extension
    assign(df_name, read.csv(file_path, stringsAsFactors = FALSE))
}

# Now, each file is a separate data frame with its own variable name

```


# Figure 1: organize data and make study site map of a) larger RI and NarrBay with Ft Weatherill and Kings Beach areas in boxes; b) zoom in of the Ft Weatherill box with specific transects as dots; c) zoom in of the Kings Beach box with specific transects as dots.
```{r}
# Create a dataframe with site coordinates
sites <- data.frame(
  Site = c("Fort Weatherill", "King's Beach"),
  Latitude = c(41.477, 41.452),  # Adjusted approximate values for the sites
  Longitude = c(-71.364, -71.326) # Adjusted approximate values for the sites
)

# Convert sites dataframe to an sf object
sites_sf <- st_as_sf(sites, coords = c("Longitude", "Latitude"), crs = 4326)

# Get Rhode Island state boundaries
ri_state <- states(cb = TRUE) %>% 
  filter(STUSPS == "RI") %>%
  st_transform(crs = 4326)

# Define the larger rectangle size (adjust as needed)
rect_size <- 0.02 

# Function to create rectangles around each site
create_rectangle <- function(site, rect_size) {
  lon <- st_coordinates(site)[1]
  lat <- st_coordinates(site)[2]
  st_as_sf(st_sfc(st_polygon(list(rbind(
    c(lon - rect_size, lat - rect_size), 
    c(lon - rect_size, lat + rect_size), 
    c(lon + rect_size, lat + rect_size), 
    c(lon + rect_size, lat - rect_size), 
    c(lon - rect_size, lat - rect_size)
  )))), crs = st_crs(site))
}

# Apply the create_rectangle function to each site
site_rectangles <- lapply(st_geometry(sites_sf), function(site) {
  create_rectangle(site, rect_size)
})

# Combine all rectangles into one sf object
site_rectangles <- do.call(rbind, site_rectangles)
st_crs(site_rectangles) <- st_crs(sites_sf) # Ensure CRS is set for combined rectangles

# Define label data
labels_tot <- data.frame(
  name = c("Narragansett\nBay", "Atlantic\nOcean"),
  lat = c(41.6, 41.3),
  lon = c(-71.35, -71.2)
)

# Plot panel A zoommed out RI and sites in rectangles
panelA <- ggplot() +
  geom_sf(data = ri_state, fill = "lightgray") + # State boundaries
  geom_sf(data = site_rectangles, color = "red2", fill = NA, size = 2) + # Add rectangles around sites
  geom_text(data = labels_tot, aes(x = lon, y = lat, label = name), size = 5, color = "black") + # Add labels
  coord_sf(xlim = c(-72, -71.0), ylim = c(41.1, 42.1), expand = FALSE) + # Zoom level
  annotation_scale(location = "bl", width_hint = 0.5) +
  annotation_north_arrow(location = "bl", which_north = "true", 
                         pad_x = unit(0.8, "in"), pad_y = unit(0.4, "in"),
                         style = north_arrow_fancy_orienteering) +
    # Change x and y axes names
  labs(x = "Longitude", y = "Latitude") +
  theme_classic() # Remove axes and background

panelA

# Save the plot to a file
ggsave("output/figures/figure_1A.png", plot = panelA, width = 8, height = 6)


# Filter transect data by site
transect_fw <- filter(site, Site == "Fort Wetherill")

# Adjust latitude values by subtracting 0.0013
transect_fw <- transect_fw %>%
  mutate(
    Start.Latitude = Start.Latitude - 0.0013
  )

# Define label data
labels <- data.frame(
  name = c("Fort\nWetherill"),
  lat = c(41.486),
  lon = c(-71.368)
)

# Plot panel B for fort wetherill transects
panelB <- ggplot() +
  geom_sf(data = ri_state, fill = "lightgray") + # State boundaries
  geom_point(data = transect_fw,
             aes(x = Start.Longitude, y = Start.Latitude),
             color = "black", fill = "red", size = 3, shape = 21) + # Add start points of transects
  annotation_scale(location = "bl", width_hint = 0.5) +
  annotation_north_arrow(location = "bl", which_north = "true", 
                         pad_x = unit(0.9, "in"), pad_y = unit(0.2, "in"),
                         style = north_arrow_fancy_orienteering) +
    geom_text(data = labels, aes(x = lon, y = lat, label = name), size = 5, color = "black") + # Add labels
  # Change x and y axes names
  labs(x = "Longitude", y = "Latitude") +
  theme_classic() + # Remove axes and background
  xlim(c(-71.45, -71.3)) + # Set longitude range
  ylim(c(41.42, 41.5))   # Set latitude range


panelB

# Save the plot to a file
ggsave("output/figures/figure_1B.png", plot = panelB, width = 8, height = 6)


# Filter transect data by site
transect_kb <- filter(site, Site == "King's Beach")

# Define label data
labels_kb <- data.frame(
  name = c("King's\nBeach"),
  lat = c(41.463),
  lon = c(-71.33)
)

# Adjust latitude values by subtracting 0.0011
transect_kb <- transect_kb %>%
  mutate(
    Start.Latitude = Start.Latitude - 0.0001
  )

# Plot panel C for kings beach transects
panelC <- ggplot() +
  geom_sf(data = ri_state, fill = "lightgray") + # State boundaries
  geom_point(data = transect_kb,
             aes(x = Start.Longitude, y = Start.Latitude),
             color = "black", fill = "red", size = 3, shape = 21) + # Add start points of transects
  annotation_scale(location = "bl", width_hint = 0.5) +
  annotation_north_arrow(location = "bl", which_north = "true", 
                         pad_x = unit(0.8, "in"), pad_y = unit(0.2, "in"),
                         style = north_arrow_fancy_orienteering) +
    geom_text(data = labels_kb, aes(x = lon, y = lat, label = name), size = 5, color = "black") + # Add labels
  # Change x and y axes names
  labs(x = "Longitude", y = "Latitude") +
  theme_classic() + # Remove axes and background
  xlim(c(-71.45, -71.3)) + # Set longitude range
  ylim(c(41.42, 41.5))   # Set latitude range


panelC

# Save the plot to a file
ggsave("output/figures/figure_1C.png", plot = panelC, width = 8, height = 6)


```

# Plot all three panels together using patchwork
```{r}

# Adjust width for individual plots
panelB <- panelB + theme(plot.margin = margin(r = 0))  # Remove right margin
panelC <- panelC + theme(plot.margin = margin(l = 0))  # Remove left margin

# Combine the plots with patchwork
combined_plot <- (panelA | (panelB / panelC)) +
  plot_annotation(tag_levels = 'A') + 
  theme(plot.tag = element_text(size = 20, face = "bold"))

# Save the combined plot
ggsave("output/figures/figure_1.png", combined_plot, width = 10, height = 6, units = "in", dpi = 300)

```


# Figure 2: organize data to make fish biomass (g/m2) on y-axis and Kelp cover (%) figure
```{r}
# take kelp cover long and just extract data for sugar kelp 2019-2023
kelp.filt <- kelp %>%
        filter(SP_CODE == "SUGK")

# Convert the integer column to character
kelp.filt$CONTROL <- as.character(kelp.filt$CONTROL)
kelp.filt$TRANSECT <- as.character(kelp.filt$TRANSECT)

# remove control column, not needed
kelp.filt <- kelp.filt[ -c(5) ]

# Read the fish aggregated CSV file 2019-2023 and skip the first column
fish <- read.csv("data/kelp_timeseries_data/fish_Aggregated.csv") %>%
          select(-1)

# Remove "KB", "FW", and space, keep only numbers
fish.filt <- fish %>%
  mutate(TRANSECT = gsub("KB |FW ", "", TRANSECT))

# Convert the integer column to character
fish.filt$TRANSECT <- as.character(fish.filt$TRANSECT)

# Rename fort weatherill
fish.filt$SITE[fish.filt$SITE == "Kings Beach"] <- "King's Beach"

# remove control column, not needed
fish.filt <- fish.filt[ -c(5) ]

# Assuming you want to join by the 'transect' column
merged.dat.cover <- left_join(fish.filt, kelp.filt)


```


# Make Figure 2 and run statistical analysis
```{r}
# Aggregate the data by year, site, and transect and compute mean values
aggregated_data_cover <- merged.dat %>%
  group_by(YEAR, SITE, TRANSECT) %>%
  summarize(mean_fish_biomass = mean(Fish.Biomass..g.m2., na.rm = TRUE),
            mean_kelp_cover = mean(COVER, na.rm = TRUE)) %>%
  ungroup()

# Filter dataframe for six top fish biomass above 150
filt_data <- aggregated_data_cover %>%
  filter(mean_fish_biomass > 150)

# Make biomass a percent so * 100
aggregated_data_cover$mean_kelp_cover <- aggregated_data_cover$mean_kelp_cover * 100

# Fit the linear mixed effects model
lmm_model <- lmer(mean_fish_biomass ~ mean_kelp_cover + SITE + (1 | SITE/YEAR), data = aggregated_data_cover)
#check normality
qqPlot(residuals(lmm_model)) #slightly non normal values, log transform to improve

# Fit the linear mixed effects model with log transformation
lmm_model <- lmer(log(mean_fish_biomass) ~ mean_kelp_cover + SITE + (1 | SITE/YEAR), data = aggregated_data_cover)
#check normality
qqPlot(residuals(lmm_model)) #improved to normal

# Summarize the model
lmm_summary <- summary(lmm_model)
lmm_tidy <- tidy(lmm_model)

# Calculate R-squared for the mixed model
r_squared <- r.squaredGLMM(lmm_model)

# Extract marginal and conditional R-squared
marginal_r2 <- round(r_squared[1], 2)
conditional_r2 <- round(r_squared[2], 2)

# Extract p-value for mean_kelp_cover
p_value_kelp_cover <- summary(lmm_model)$coefficients["mean_kelp_cover", "Pr(>|t|)"]

# Create scatter plot
figure.2 <- ggplot(aggregated_data_cover, aes(x = mean_kelp_cover, y = mean_fish_biomass)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, formula = y ~ x, color = "blue") +
annotate("text", x = Inf, y = Inf, label = paste0("Marginal R^2 = ", marginal_r2, "\nConditional R^2 = ", conditional_r2, "\np = ", signif(p_value_kelp_cover, 3)), hjust = 1.4, vjust = 1.1, size = 4, parse = FALSE) + # Adjust vjust
  labs(x = "Kelp cover (%)", y = "Fish biomass (g/m2)",
       title = "Fish Biomass vs Kelp Cover") +
  theme_classic();figure.2

# Save the plot to a file
ggsave("output/figures/figure_2.png", plot = figure.2, width = 8, height = 6)
```


# Figure 3: organize and calculate kelp biomass data to make fish biomass/kelp biomass (g/m2) linear regression figure
```{r}
# read in quad data with filtered for just kelp
quad.filt <- QUAD %>%
        filter(SP_CODE == "SUGK")

# remove control column and two last columns, not needed
quad.filt <- quad.filt[ -c(13:14) ] 
quad.filt <- quad.filt[ -c(8) ] 
quad.filt <- quad.filt[ -c(5:6) ] 

# I noticed that 06/30/23 dates were accidentily labeled 06/30/29
# Replace "29" with "23" in the DATE column
quad.filt$DATE <- sub("29$", "23", quad.filt$DATE)

# Convert DATE to Date format
quad.filt$DATE <- as.Date(quad.filt$DATE, format = "%m/%d/%y")

# Convert Date format to "YYYY-MM-DD"
quad.filt$DATE <- format(quad.filt$DATE, "%Y-%m-%d")

# combine quad data with morph data to calculate kelp biomass

#load in morph data and select for SUGK
morph.filt <- morph %>%
        filter(SP_CODE == "SL")

#rename SL to SUGK
morph.filt$SP_CODE[morph.filt$SP_CODE == "SL"] <- "SUGK"

# Convert the integer column to character
morph.filt$TRANSECT <- as.character(morph.filt$TRANSECT)

# merge morph filt and quad.filt to calculate biomass
merge.morph.quad <- left_join(morph.filt, quad.filt)

#calculate kelp biomass
# To get wet weight in grams (biomass) you use the equation: y=0.00421*x^1.951 where x is kelp blade + stipe length.
# Calculate the sum of blade length and stipe length
merge.morph.quad$sum_length <- merge.morph.quad$BLADE_LENGTH_CM + merge.morph.quad$STIPE_LENGTH_CM

# Define a function to calculate kelp biomass
calculate_kelp_biomass <- function(x) {
  y <- 0.00421 * x^1.951
  return(y)
}

# Calculate kelp biomass using the sum of blade length and stipe length
merge.morph.quad$kelp_biomass <- calculate_kelp_biomass(merge.morph.quad$sum_length)

# Remove the intermediate columns if no longer needed
merge.morph.quad$kelp_blade_length <- NULL
merge.morph.quad$sum_length <- NULL

# Replace NA values with 0
merge.morph.quad$kelp_biomass <- ifelse(is.na(merge.morph.quad$kelp_biomass), 0, merge.morph.quad$kelp_biomass)

# Sum kelp biomass at the transect level
morph.filt <- merge.morph.quad %>%
  group_by(TRANSECT, YEAR, SITE) %>%
  summarize(total_kelp_biomass = sum(kelp_biomass))

# Divide by 6 quadrats for correct g/m2 calculation
morph.filt$average_kelp_biomass <- morph.filt$total_kelp_biomass / 6

# Merge fish and morph data
merged.dat.morph <- left_join(morph.filt, fish.filt)

```


# Make figure 3 and run statistical analysis
```{r}
# Aggregate the data by year, site, and transect and compute mean values
aggregated.data.morph <- merged.dat.morph %>%
  group_by(YEAR, SITE, TRANSECT) %>%
  summarize(mean_fish_biomass = mean(Fish.Biomass..g.m2., na.rm = TRUE),
            mean_kelp_biomass = mean(average_kelp_biomass, na.rm = TRUE)) %>%
  ungroup()

# Fit the linear mixed effects model
lmm_model <- lmer(mean_fish_biomass ~ mean_kelp_biomass + SITE + (1 | SITE/YEAR), data = aggregated.data.morph)
#check normality
qqPlot(residuals(lmm_model)) #slightly non normal values, log transform to improve

# Fit the linear mixed effects model with log transformation
lmm_model <- lmer(log(mean_fish_biomass) ~ mean_kelp_biomass + SITE + (1 | SITE/YEAR), data = aggregated.data.morph)
#check normality
qqPlot(residuals(lmm_model)) #improved to normal

# Summarize the model
lmm_summary <- summary(lmm_model)
lmm_tidy <- tidy(lmm_model)

# Calculate R-squared for the mixed model
r_squared <- r.squaredGLMM(lmm_model)

# Extract marginal and conditional R-squared
marginal_r2 <- round(r_squared[1], 2)
conditional_r2 <- round(r_squared[2], 2)

# Extract p-value for mean_kelp_cover
p_value_kelp_biomass <- summary(lmm_model)$coefficients["mean_kelp_biomass", "Pr(>|t|)"]

# Create scatter plot
figure.3 <- ggplot(aggregated.data.morph, aes(x = mean_kelp_biomass, y = mean_fish_biomass)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, formula = y ~ x, color = "blue") +
annotate("text", x = Inf, y = Inf, label = paste0("Marginal R^2 = ", marginal_r2, "\nConditional R^2 = ", conditional_r2, "\np = ", signif(p_value_kelp_biomass, 3)), hjust = 1.4, vjust = 1.1, size = 4, parse = FALSE) + # Adjust vjust
  labs(x = "Kelp Biomass (g/m2)", y = "Fish biomass (g/m2)",
       title = "Fish Biomass vs Kelp Biomass") +
  theme_classic();figure.3

# Save the plot to a file
ggsave("output/figures/figure_3.png", plot = figure.3, width = 8, height = 6)
```


# Figure 4: organize rugosity data and make linear regression figure for fish biomass and rugosity
```{r}
# Calculate mean rugosity complexity scores across depths for each transect, grouped by site and year
mean_rugosity <- rugosity %>%
  group_by(SITE, YEAR, TRANSECT) %>%
  summarize(mean_rugosity = mean(c(X0M, X4M, X8M, X12M, X16M, X20M, X24M, X28M, X32M, X36M, X40M), na.rm = TRUE))

# change transect from integer to chr
mean_rugosity$TRANSECT <- as.character(mean_rugosity$TRANSECT)

# Left jon fish and rugosity data
merged.dat.rug <- left_join(mean_rugosity, fish.filt)

```


# Make figure 4 and run statistical analysis
```{r}
# Aggregate the data by year, site, and transect and compute mean values
aggregated.data.rug <- merged.dat.rug %>%
  group_by(YEAR, SITE, TRANSECT) %>%
  summarize(mean_fish_biomass = mean(Fish.Biomass..g.m2., na.rm = TRUE),
            mean_rugosity = mean(mean_rugosity, na.rm = TRUE)) %>%
  ungroup()

# Fit the linear mixed effects model
lmm_model <- lmer(mean_fish_biomass ~ mean_rugosity + SITE + (1 | SITE/YEAR), data = aggregated.data.rug)
#check normality
qqPlot(residuals(lmm_model)) #data is normal no need to transform

# Summarize the model
lmm_summary <- summary(lmm_model)
lmm_tidy <- tidy(lmm_model)

# Calculate R-squared for the mixed model
r_squared <- r.squaredGLMM(lmm_model)

# Extract marginal and conditional R-squared
marginal_r2 <- round(r_squared[1], 2)
conditional_r2 <- round(r_squared[2], 2)

# Extract p-value for mean_kelp_cover
p_value_rugosity <- summary(lmm_model)$coefficients["mean_rugosity", "Pr(>|t|)"]

# Create scatter plot
figure.4 <- ggplot(aggregated.data.rug, aes(x = mean_rugosity, y = mean_fish_biomass)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, formula = y ~ x, color = "blue") +
  annotate("text", x = Inf, y = Inf, label = paste0("Marginal R^2 = ", marginal_r2, "\nConditional R^2 = ", conditional_r2, "\np = ", signif(p_value_rugosity, 2)), 
           hjust = 2.1, vjust = 1.1, size = 4, parse = FALSE) +
  labs(x = "Rugosity Complexity", y = "Fish biomass (g/m2)",
       title = "Fish Biomass vs Rugosity Complexity") +
  theme_classic();figure.4

# Save the plot to a file
ggsave("output/figures/figure_4.png", plot = figure.4, width = 8, height = 6)
```




