---
title: "manuscript_figures"
author: "daniellembecker"
date: "2024-05-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Use this script to use cleaned data and make plots for RI kelp data manuscript

# Load libraries 
```{r}
library(plyr)
library(dplyr)
library(MuMIn)
library(lmerTest)
library(tidyverse)
library(lme4)
library(broom.mixed)
library(lubridate)
library(ggplot2)
library(patchwork)
library(broom)

```

# load cleaned dataframes

```{r}
# Set your directory path
directory <- "output/cleaned_data/"

# List files in the directory
files <- list.files(directory)

# Initialize an empty list to store data frames
data_frames <- list()

# Loop through each file
for (file in files) {
    # Read the file into a data frame
    # Adjust the read function according to your file type (e.g., read.csv, read.table)
    # Specify other parameters as needed (e.g., header = TRUE)
    file_path <- file.path(directory, file)
    df_name <- gsub("\\..*", "", file)  # Extract file name without extension
    assign(df_name, read.csv(file_path, stringsAsFactors = FALSE))
}

# Now, each file is a separate data frame with its own variable name

```

# Figure 2: Fish biomass (g/m2) on y-axis and Kelp cover (%) on the x-axis. Point scatter plot with a linear regression and 95% CI bands. For all data points, you will have one value for fish biomass and a corresponding value for kelp cover. So it will be by: year x site x transect.

```{r}
# take kelp cover long and just extract data for sugar kelp 2019-2023
kelp.filt <- kelp %>%
        filter(SP_CODE == "SUGK")

# Convert the integer column to character
kelp.filt$CONTROL <- as.character(kelp.filt$CONTROL)
kelp.filt$TRANSECT <- as.character(kelp.filt$TRANSECT)

# remove control column, not needed
kelp.filt <- kelp.filt[ -c(5) ]

# Read the fish aggregated CSV file 2019-2023 and skip the first column
fish <- read.csv("data/kelp_timeseries_data/fish_Aggregated.csv") %>%
          select(-1)

# Remove "KB", "FW", and space, keep only numbers
fish.filt <- fish %>%
  mutate(TRANSECT = gsub("KB |FW ", "", TRANSECT))

# Convert the integer column to character
fish.filt$TRANSECT <- as.character(fish.filt$TRANSECT)

# Rename fort weatherill
fish.filt$SITE[fish.filt$SITE == "Kings Beach"] <- "King's Beach"

# remove control column, not needed
fish.filt <- fish.filt[ -c(5) ]

# Assuming you want to join by the 'transect' column
merged.dat <- left_join(fish.filt, kelp.filt)


```


# Make Figure 2
```{r}

# Aggregate the data by year, site, and transect and compute mean values
aggregated_data <- merged.dat %>%
  group_by(YEAR, SITE, TRANSECT) %>%
  summarize(mean_fish_biomass = mean(Fish.Biomass..g.m2., na.rm = TRUE),
            mean_kelp_cover = mean(COVER, na.rm = TRUE)) %>%
  ungroup()

# Fit the linear mixed effects model
lmm_model <- lmer(mean_fish_biomass ~ mean_kelp_cover + SITE + (1 | SITE/YEAR), data = aggregated_data)

# Summarize the model
lmm_summary <- summary(lmm_model)
lmm_tidy <- tidy(lmm_model)

# Calculate R-squared for the mixed model
r_squared <- r.squaredGLMM(lmm_model)

# Extract marginal and conditional R-squared
marginal_r2 <- round(r_squared[1], 2)
conditional_r2 <- round(r_squared[2], 2)

# Extract p-value for mean_kelp_cover
p_value_kelp_cover <- summary(lmm_model)$coefficients["mean_kelp_cover", "Pr(>|t|)"]

# Create scatter plot
figure.2 <- ggplot(aggregated_data, aes(x = mean_kelp_cover, y = mean_fish_biomass)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, formula = y ~ x, color = "blue") +
annotate("text", x = Inf, y = Inf, label = paste0("Marginal R^2 = ", marginal_r2, "\nConditional R^2 = ", conditional_r2, "\np = ", signif(p_value_kelp_cover, 3)), hjust = 1.4, vjust = 1.1, size = 4, parse = FALSE) + # Adjust vjust
  labs(x = "Kelp cover (%)", y = "Fish biomass (g/m2)",
       title = "Fish Biomass vs Kelp Cover") +
  theme_classic();figure.2

# Save the plot to a file
ggsave("output/figures/figure_2.png", plot = figure.2, width = 8, height = 6)
```


# Figure 3: Make the saame regression figure for fish biomass and kelp biomass (g/m2). To get kelp biomass, you will use the quad data which counts the number of kelp blades, and convert to biomass using the morphometric data. Take stipe length, blade width, blade length and convert to biomass. Then get a mean and SE of kelp frond biomass at that transect and multiply that by the number of fronds counted in quad. We want units to be g/m2 (I think). 









